Modelling the creation of ecological traps by environmentally persistent pathogens

### 2015-05-27 Manuscript/manuscript.tex
More discussion edits (relative benefit of high v. low)

@@

Because high quality habitat supports larger host population sizes, host populations on high quality habitat are less susceptible to environmental stochasticity and thus experience fewer extinction events.
Due to this additional stability, high quality habitat is more consistently occupied, by both susceptible and infectious hosts, leading to the increase in infection probability with patch quality.
From the perspective of a migrant susceptible host then, selection of high quality habitat leads to increased disease risk relative to other habitats, either because the patch is likely already occupied by infected hosts, or because susceptible hosts are likely to become infected (either through the environmental reservoir or direct contact).
Thus, when high quality habitat is preferentially selected by migrants (i.e. $\xi_em > 0$), high quality habitats form ecological traps, that is, preferentially selected habitat that reduces fitness relative to other habitats \cite{Robertson2006}).
Moreover, this preferential selection of high quality habitat further exacerbates the increase in disease risk with habitat quality (Fig~\ref{pinfection}) by increasing the risk of direct infection from infectious migrants also selecting high quality habitat, reflecting results from the contact network literature that show that highly connected nodes have higher infection risk (\cite{Christley2005}, \cite{Keeling2005}).

@@

In contrast to high quality habitat, low quality patches help to limit pathogen spread and increase susceptible occupancy (Fig~\ref{endemic}, \ref{popsizes}). 
Even though individual low quality patches, due to their high extinction rates and low colonization rates, are unable to support consistent occupancy (Fig~\ref{simvis}), the presence of low quality patches facilitates susceptible persistence for a greater range of pathogens (Fig~\ref{endemic}).  
This phenomenon represents the other side of the role played by high quality patches in that low quality patches are relatively unstable (i.e. low population sizes and high extinction rates) and thus do not support infectious populations for long.
Since they are infrequently colonized (when $\xi_em > 0$), the environmental reservoir left behind by these infectious populations likely decays before it has the opportunity to infect new susceptible colonists.  
As a result, low quality patches effectively represent a dead-end for the pathogen, reducing the number of patches from which it can spread, and thereby reducing infectious occupancy (Fig~\ref{endemic}, \ref{popsizes}).

Though there is a consistent trend in the effect of low and high quality habitat on pathogen spread, the net consequences of these effects for total occupancy vary depending on the traits of the pathogen.  Specifically, the findings that low quality habitat inhibits disease spread while high quality habitat enhances it suggest the presence of a trade-off between between processes that facilitate metapopulation stability at the expense of disease spread, and processes that hinder disease spread at the expense of metapopulation stability.  This work thus expands on previous studies of disease spread in a metapopulation that focused on movement rate as the driver of this trade-off, with increased movement both improving metapopulation stability and increasing disease spread (\cite{Hess1996}, \cite{Gog2002}, \cite{Park2012}).  Shifting the patch quality distribution towards high quality patches in our model is in many ways analogous to increasing movement in the models of Hess, Gog, and Park, but by including habitat heterogeneity and metapopulation structure, we are able to investigate the role of individual patches and their contribution to total occupancy.

These contributions, and the relative strength of the above trade-offs, depend largely on the role of connectivity, how it is affected by patch quality, and the resulting strength and structure of the different avenues of pathogen transmission.  Broadly, we see three different scenarios that roughly correspond to the three sections of the "u"-shaped relationships in Fig~\ref{highvlow}: conditions that support very weak pathogen spread, conditions that facilitate intermediate/endemic pathogen spread, and conditions that produce easy/pandemic pathogen spread.  When pathogen spread is weak (e.g. due to low longevity; the top left of the "u"), high quality patches are a net benefit to the metapopulation, as their connectivity and stability facilitate greater occupancy with little additional pathogen spread.

However, in cases where the pathogen is able to spread more easily (i.e. at an endemic level; the trough of the "u"), the addition of high quality patches functions similarly to increasing movement in the model developed by \cite{Hess1996}: increasing pathogen spread and as a consequence reducing overall occupancy (Fig~\ref{sens}c).  In these cases, the reduction in occupancy from the effect of high quality habitat on pathogen spread outweighs the positive consequences of increased metapopulation connectivity and stability.  The shift in this trade-off can arise from conditions that change either side of that balance.  High quality habitat distributions require lower pathogen longevities for pathogen spread to begin to take off, relative to low quality habitat distributions.  Thus, for intermediate longevities, high quality distributions experience considerably greater pathogen spread and lower occupancy than low quality distributions.

>>>>

Blah blah blah....something about lattice looking similar (supp info).

@@

Because high quality habitat supports larger host population sizes, host populations on high quality habitat are less susceptible to environmental stochasticity and thus experience fewer extinction events.
Due to this additional stability, high quality habitat is more consistently occupied, by both susceptible and infectious hosts, leading to the increase in infection probability with patch quality.
From the perspective of a migrant susceptible host then, selection of high quality habitat leads to increased disease risk relative to other habitats, either because the patch is likely already occupied by infected hosts, or because susceptible hosts are likely to become infected (either through the environmental reservoir or direct contact).
Thus, when high quality habitat is preferentially selected by migrants (i.e. $\xi_{em} > 0$), high quality habitats form ecological traps, that is, preferentially selected habitat that reduces fitness relative to other habitats \cite{Robertson2006}).
Moreover, this preferential selection of high quality habitat further exacerbates the increase in disease risk with habitat quality (Fig~\ref{pinfection}) by increasing the risk of direct infection from infectious migrants also selecting high quality habitat, reflecting results from the contact network literature that show that highly connected nodes have higher infection risk (\cite{Christley2005}, \cite{Keeling2005}).

@@

In contrast to high quality habitat, low quality patches help to limit pathogen spread and increase susceptible occupancy (Fig~\ref{endemic}, \ref{popsizes}). 
Even though individual low quality patches, due to their high extinction rates and low colonization rates, are unable to support consistent occupancy (Fig~\ref{simvis}), the presence of low quality patches facilitates susceptible persistence for a greater range of pathogens (Fig~\ref{endemic}).  
This phenomenon represents the other side of the role played by high quality patches in that low quality patches are relatively unstable (i.e. low population sizes and high extinction rates) and thus do not support infectious populations for long.
Since they are infrequently colonized (when $\xi_{em} > 0$), the environmental reservoir left behind by these infectious populations likely decays before it has the opportunity to infect new susceptible colonists.  
As a result, low quality patches effectively represent a dead-end for the pathogen, reducing the number of patches from which it can spread, and thereby reducing infectious occupancy (Fig~\ref{endemic}, \ref{popsizes}).

These findings that low quality habitat inhibits disease spread while high quality habitat enhances it suggest the presence of a trade-off between processes that facilitate metapopulation stability at the expense of disease spread, and processes that hinder disease spread at the expense of metapopulation stability.
In this light, this work expands on previous studies of disease spread in a metapopulation that focused on movement rate as the driver of this trade-off, with increased movement both improving metapopulation stability and increasing disease spread (\cite{Hess1996}, \cite{Gog2002}, \cite{Park2012}).  
Shifting the patch quality distribution towards high quality habitat in our model is  conceptually similar to increasing movement in these models, and as in these models, the net consequences of this trade-off for the host metapopulation depend on the traits of the pathogen.
Broadly, we see three different scenarios: pathogen traits produce very weak pathogen spread and high quality habitat is beneficial; pathogen traits facilitate intermediate/endemic pathogen spread and low quality habitat is beneficial; and pathogen traits produce easy/pandemic pathogen spread and high quality habitat is again beneficial.

When pathogen spread is weak, high quality patches are a net benefit to the metapopulation, as their connectivity and stability facilitate greater occupancy with little additional pathogen spread.

However, in cases where the pathogen is able to spread more easily (i.e. at an endemic level), the addition of high quality patches functions similarly to increasing movement in the model developed by \cite{Hess1996}: increasing pathogen spread and as a consequence reducing overall occupancy (Fig~\ref{sens}c).  In these cases, the reduction in occupancy from the effect of high quality habitat on pathogen spread outweighs the positive consequences of increased metapopulation connectivity and stability.  The shift in this trade-off can arise from conditions that change either side of that balance.  High quality habitat distributions require lower pathogen longevities for pathogen spread to begin to take off, relative to low quality habitat distributions.  Thus, for intermediate longevities, high quality distributions experience considerably greater pathogen spread and lower occupancy than low quality distributions.
